---
title: "Assignment_6_HPC"
author: "Cameron DeBellefeuille"
date: "2025-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Github

[Github](https://github.com/CameronDebellefeuille/BIOL432_Assignment6_HPC)

## 2. New R Markdown

Complete.

## 3. Libraries

```{r, warning=F}
library(doParallel)
library(parallel)
library(foreach)
```

## 4. detectCores()

```{r}
cores <- detectCores()
```

The output of the detectCores() functions indicates my computer has **8 cores**.

## 5. Creating a loop and 6. Tracking the time

```{r}
Start <- Sys.time()
for (i in 1:4000){
  mean(rnorm(100000, mean = 10, sd = 3))
}
SerTime <- Sys.time() - Start
SerTime
```

## 7. Predicting time improvement with multiple threads

```{r}
PredTime <- SerTime / cores # divide the serial loop time by number of cores
PredTime
```

```{r}
pred_timediff <- SerTime - PredTime # Expected time saved
pred_timediff
```

Since my system has **8 cores**, the predicted runtime is **approximately 1/8th of the serial execution time**. The estimated time improvement is around **29.8 seconds**.

## 8. Running the loop in parallel and 9. Calculating new run time

```{r}
# Setting up
cores <- detectCores() - 1 # Using 7 cores to avoid system overload
cluster <- makeCluster(cores)
registerDoParallel(cluster)

# Running loop
Start2 <- Sys.time()
results <- foreach(i = 1:4000, .combine=rbind) %dopar% {
  x <- mean(rnorm(100000, mean = 10, sd = 3))
}
ParaTime <- Sys.time() - Start2

# Freeing up cores
stopCluster(cluster)

print(ParaTime)
```

## 10. Comparing serial loop time vs parallel loop time

```{r}
time_saved <- SerTime - ParaTime # Actual time saved
print(time_saved)
```

```{r}
ParaTime - PredTime
```

The parallel loop was approximately **21.4 seconds faster** than the serial loop. However, the actual parallel execution time was **8.5 seconds longer** than the theoretical prediction. This discrepancy is due to the initializing of clusters, the printing of iteration results, and the utilization of only 7 out of 8 cores. Most importantly however, **data merging** (combine = rbind) added extra computation time, as it was necessary to merge results from parallel loops, leading to a slower-than-expected runtime.
